---
title: "CARGO_CLAIMS"
format: html
editor_options: 
  chunk_output_type: console
---

## Reading in the data obviiiii
```{r}

pacman::p_load(tidyverse, readxl, ggrepel, directlabels, readr, rio, knitr, pander, downloader, dplyr, plotly, ggplot2, ggthemes, haven, ggdist, janitor, MASS, here)

path <- "srcsc-2026-claims-cargo.xlsx"

cargo_freq <- read_excel(path, sheet = "freq")
cargo_sev <- read_excel(path, sheet = "sev")


```

## Combing Rows and Filtering cargo_freq

```{r}

cargo_pol <- cargo_freq %>%
  mutate(
    claim_count = as.numeric(claim_count),
    exposure    = as.numeric(exposure)
  ) %>%
  filter(
    is.finite(claim_count),
    claim_count >= 0,      # <- key addition
    is.finite(exposure),
    exposure > 0
  ) %>%
  group_by(policy_id) %>%
  summarise(
    exposure_pol = sum(exposure),
    claims_pol   = sum(claim_count),
    .groups = "drop"
  ) %>%
  filter(exposure_pol > 0)


```

## Estimating claim Frequency

```{r}

# Find your exposure - adjusted claim rate first AKA the expected number of claims per unit of exposure. This is NOT the average number of claims per policy since all exposures are NOT exactly equal to one.

lambda_hat <- sum(cargo_pol$claims_pol) / sum(cargo_pol$exposure_pol)
lambda_hat

# Now find the average number of claims per policy with NO exposure adjustment. Also find the variance as well.

mean_claims <- mean(cargo_pol$claims_pol)
var_claims  <- var(cargo_pol$claims_pol)

# Using your mean and variance now assign a distribution to the claim counts. It will either be Poisson or Negative Binomial. YOU WILL NOT USE THE LAMBDA WHEN ASSIGNING THE DISTRIBUTION. This is because the distribution choice is about shape, not scale, so you must use raw observed counts NOT exposure-scaled rates.

dispersion_ratio <- var_claims / mean_claims
c(mean_claims = mean_claims, var_claims = var_claims, dispersion_ratio = dispersion_ratio)

## this dispersion ratio shows how much the actual variance exceeds Poisson Variance. For dispersion ratios if its 0.8 - 1.2, then Poisson is a good bet. If it's anything greater then Negative Binomial is your next best bet! 


```

## Moving on to Serverity Modeling (this just means the size of the claims)

```{r}

## In this chunk I'm cleaning the data a bit, making sure I only use positive entries. For severity modeling in actuarial science typically you will use either gamma or log normal. I will start with log normal and then later compare it to Gamma using the MOM. 

x <- cargo_sev$claim_amount
x <- as.numeric(x)
x <- x[is.finite(x) & !is.na(x) & x > 0]

# Finding my distribution parameters

mu    <- mean(log(x))
sigma <- sd(log(x))

mean_sev   <- exp(mu + 0.5*sigma^2)
median_sev <- exp(mu)

c(meanlog = mu, sdlog = sigma, mean_severity = mean_sev, median_severity = median_sev)

```

## Visuals
```{r}

# A histogram of my log normal.

hist(
  log(x),
  breaks = 50,
  main = "Log of Cargo Claim Amounts",
  xlab = "log(Claim Amount)"
)

p<- qqplot(
  qlnorm(ppoints(length(x)), meanlog = mu, sdlog = sigma),
  sort(x),
  main = "QQ Plot – Cargo Severity (Lognormal)",
  xlab = "Theoretical Quantiles",
  ylab = "Observed Claim Amounts"
) 
abline(0, 1, col = "red", lwd = 2)

cargo_sev$cargo_type_clean <- sub("_.*", "", cargo_sev$cargo_type)

ggplot(cargo_sev, aes(x = log(claim_amount), fill = cargo_type_clean)) +
  geom_density(alpha = 0.4) +
  labs(title = "Log Severity by Cargo Type")

```

## Comapring Distribution to Gamma

```{r}

# Just so you know AIC is a model comparison test that uses the likelihood of the fitted model you chose and ranks the models.

# Lognormal log-likelihood + AIC 
loglik_logn <- sum(dlnorm(x, meanlog = mu, sdlog = sigma, log = TRUE))
aic_logn <- 2*2 - 2*loglik_logn

# Gamma MOM estimates + log-likelihood + AIC 
m <- mean(x); v <- var(x)
shape_hat <- m^2 / v
rate_hat  <- m / v

loglik_gam <- sum(dgamma(x, shape = shape_hat, rate = rate_hat, log = TRUE))
aic_gam <- 2*2 - 2*loglik_gam

data.frame(
  Model = c("Lognormal", "Gamma"),
  AIC   = c(aic_logn, aic_gam),
  row.names = NULL
)

# CHOOSE THE DISTRIBUTION WITH THE LOWER AIC!!!!!!

```

## Tail Summaries

```{r}

q95 <- qlnorm(0.95, meanlog = mu, sdlog = sigma)
q99 <- qlnorm(0.99, meanlog = mu, sdlog = sigma)

c(q95_logn = q95, q99_logn = q99)

quantile(x, probs = c(0.90, 0.95, 0.99))
mean(x)

```

## Final report Sumamry
```{r}

# I had chat write this up for me. You guys can jsut copy and paste this with your appropriate variable names. 

delta_aic <- aic_gam - aic_logn

cat(
  "- **Baseline frequency (lambda_hat):** ", round(lambda_hat, 4), " claims per exposure-year\n",
  "- **Overdispersion (Var/Mean):** ", round(dispersion_ratio, 3), " → Negative Binomial preferred\n",
  "- **Severity model:** Lognormal (meanlog = ", round(mu, 3), ", sdlog = ", round(sigma, 3), ")\n",
  "- **Mean severity (Lognormal):** $", format(round(mean_sev, 0), big.mark=","), "\n",
  "- **AIC comparison (Gamma - Lognormal):** ΔAIC = ", round(delta_aic, 1), " (positive favors Lognormal)\n",
  sep = ""
)

```



## PART TWOOOOOOO

```{r}

# Start by getting parameters for the distribution you chose for your claim freq count. (Either Poisson or NB) I will be doing it for NB since that's what i chose in Part 1. 

# Finding my parameters while taking into account exposure!! NegBin with exposure offset: E[claims | exposure] = rate * exposure. Also just as a side note in the traditional NB we use (r,p), in this case we are setting r = theta and p = theta/(theta+mu). This is common practice in actuarial pricing. 
fit_nb <- glm.nb(claims_pol ~ 1 + offset(log(exposure_pol)), data = cargo_pol)
summary(fit_nb)

# After you fit the model you can now find the actual parameters.
lambda_hat_nb <- exp(coef(fit_nb)[1])  
# lamba is rate per exposure unit
theta_nb      <- fit_nb$theta           
#  theta is the dispersion parameter AKA our (p) parameter

c(lambda_hat_nb = lambda_hat_nb, theta_nb = theta_nb)

```

## Time to use your severity and freq distributions to run some simulations!! We are using these simulations to see what the expected losses will be on average in the worst stress years as well as the worst 1% of years. (This is what Var and TVaR is)

```{r}

# you must start by setting your exposure period which is the sum of your exposure

total_exposure <- sum(cargo_pol$exposure_pol)

Exposure_period <- total_exposure 

mu_period <- as.numeric(lambda_hat_nb) * Exposure_period
mu_period

# this is the expected/theoretical mean that we are shooting for. We will use this number to compare to the simulated mean

expected_aggregate <- as.numeric(lambda_hat_nb) * Exposure_period * mean_sev
expected_aggregate

# SIMULATION TIME YAYYYY - i stole bootstrapping code from 424

set.seed(123)
n_sims <- 30000

mu_period <- as.numeric(lambda_hat_nb) * Exposure_period

agg_losses <- numeric(n_sims)

for (i in seq_len(n_sims)) {
  N <- rnbinom(1, size = theta_nb, mu = mu_period)
  if (N == 0) {
    agg_losses[i] <- 0
  } else {
    agg_losses[i] <- sum(rlnorm(N, meanlog = mu, sdlog = sigma))
  }
}

# these are the values we will be putting in a table that we want

mean_L  <- mean(agg_losses)
VaR_99  <- as.numeric(quantile(agg_losses, 0.99, names = FALSE))
TVaR_99 <- mean(agg_losses[agg_losses >= VaR_99])

# in the table below you'll notice my ratio is super close to 1 this is AWESOME and means the simulation is correct.

c(sim_mean = mean_L, theo_mean = expected_aggregate, ratio = mean_L/expected_aggregate,
  VaR_99 = VaR_99, TVaR_99 = TVaR_99)

```

## Fitting models/simulation for each solar system

```{r}

# find the biggest risks from each solar system that DIRECTLY effect your specific hazard. For example for Cargo debris density obvi effects cargo collisions, so when reading about each system I saw that Heli had the most debris so i put it in the .95 quantile compared to the other solar systems.


#Also on the lowest of keys i gave chat info about each of the solar systems and 

q <- function(v, p) as.numeric(quantile(v, p, na.rm = TRUE))

profiles <- rbind(
  Helionis = data.frame(
    solar_radiation  = q(cargo_pol$solar_radiation, 0.25),
    debris_density   = q(cargo_pol$debris_density, 0.95),
    route_risk       = q(cargo_pol$route_risk, 0.85),
    distance         = q(cargo_pol$distance, 0.50),
    transit_duration = q(cargo_pol$transit_duration, 0.50),
    exposure_pol     = 1
  ),
  Bayesia = data.frame(
    solar_radiation  = q(cargo_pol$solar_radiation, 0.90),
    debris_density   = q(cargo_pol$debris_density, 0.40),
    route_risk       = q(cargo_pol$route_risk, 0.40),
    distance         = q(cargo_pol$distance, 0.60),
    transit_duration = q(cargo_pol$transit_duration, 0.60),
    exposure_pol     = 1
  ),
  Oryn = data.frame(
    solar_radiation  = q(cargo_pol$solar_radiation, 0.60),
    debris_density   = q(cargo_pol$debris_density, 0.50),
    route_risk       = q(cargo_pol$route_risk, 0.75),
    distance         = q(cargo_pol$distance, 0.90),
    transit_duration = q(cargo_pol$transit_duration, 0.90),
    exposure_pol     = 1
  )
)

profiles

```

```{r}
# baseline profile = "typical" environment
baseline <- data.frame(
  solar_radiation  = median(cargo_pol$solar_radiation, na.rm=TRUE),
  debris_density   = median(cargo_pol$debris_density, na.rm=TRUE),
  route_risk       = median(cargo_pol$route_risk, na.rm=TRUE),
  distance         = median(cargo_pol$distance, na.rm=TRUE),
  transit_duration = median(cargo_pol$transit_duration, na.rm=TRUE),
  exposure_pol     = 1
)

# Frequency (claims per exposure unit)
freq_base <- as.numeric(predict(fit_nb_x, newdata = baseline,  type = "response"))
freq_sys  <- as.numeric(predict(fit_nb_x, newdata = profiles, type = "response"))

freq_mult <- freq_sys / freq_base
names(freq_mult) <- rownames(profiles)
freq_mult

mode_type <- names(sort(table(cargo_sev_clean$cargo_type_clean), decreasing = TRUE))[1]

baseline_sev <- transform(baseline, cargo_type_clean = mode_type)
profiles_sev <- cbind(profiles, cargo_type_clean = mode_type)

m_base <- as.numeric(predict(fit_sev, newdata = baseline_sev))
m_sys  <- as.numeric(predict(fit_sev, newdata = profiles_sev))

sev_mult <- exp(m_sys - m_base)
names(sev_mult) <- rownames(profiles)
sev_mult


```

```{r}

simulate_system <- function(E_period, freq_mult, sev_mult, n_sims = 30000, seed = 123) {
  set.seed(seed)
  mu_period <- as.numeric(lambda_hat_nb) * freq_mult * E_period

  agg <- numeric(n_sims)
  for (i in seq_len(n_sims)) {
    N <- rnbinom(1, size = theta_nb, mu = mu_period)
    agg[i] <- if (N == 0) 0 else sum(rlnorm(N, meanlog = mu + log(sev_mult), sdlog = sigma))
  }

  VaR_99  <- as.numeric(quantile(agg, 0.99, names = FALSE))
  TVaR_99 <- mean(agg[agg >= VaR_99])

  c(E_L = mean(agg), Var_L = var(agg), VaR_99 = VaR_99, TVaR_99 = TVaR_99)
}

E_total <- total_exposure 

# simulating for each system based off profiles made earlier

out <- rbind(
  Helionis = simulate_system(E_total, freq_mult["Helionis"], sev_mult["Helionis"]),
  Bayesia  = simulate_system(E_total, freq_mult["Bayesia"],  sev_mult["Bayesia"]),
  Oryn     = simulate_system(E_total, freq_mult["Oryn"],     sev_mult["Oryn"])
)

#this is the end goal table we want 
out


```



