---
title: "workers compensation"
format: html
editor_options: 
  chunk_output_type: console
---

## Load data - libraries

```{r}

pacman::p_load(tidyverse, readxl, ggrepel, directlabels, readr, rio, knitr, pander, downloader, dplyr, plotly, ggplot2, ggthemes, haven, ggdist, janitor, MASS, here)

path = "srcsc-2026-claims-workers-comp.xlsx"

work_comp_freq = read_excel(path, sheet = "freq")
work_comp_sev = read_excel(path, sheet = "sev")


```


## Combing Rows and Filtering

```{r}

work_comp_pol = work_comp_freq %>%
  filter(!is.na(exposure), !is.na(claim_count), exposure > 0) %>%
  group_by(policy_id) %>%
  summarise(
    exposure_pol = sum(exposure, na.rm = TRUE),
    claims_pol   = sum(claim_count, na.rm = TRUE),
    .groups = "drop"
  )

```

## Estimating claim Frequency

```{r}

# Expected number of claims, normalized by exposure

lambda_hat = sum(work_comp_pol$claims_pol) / sum(work_comp_pol$exposure_pol)
lambda_hat

# expected number of claims per policy with NO exposure adjustment.

mean_claims = mean(work_comp_pol$claims_pol)
var_claims  = var(work_comp_pol$claims_pol)

# variance normalized by expected claims per policy (dispersion)

dispersion_ratio = var_claims / mean_claims
c(mean_claims = mean_claims, var_claims = var_claims, dispersion_ratio = dispersion_ratio)

# we use dispersion to see if possion would be a good fit (mean = vairance)
# we will use neg-binomial, as varaince is 50% greater than mean

```

## Moving on to Serverity Modeling (this just means the size of the claims)

```{r}

# grab the claim amount $$, and filter it

x = work_comp_sev$claim_amount
x = as.numeric(x)
x = x[is.finite(x) & !is.na(x) & x > 0]

# distribution parameters for claim amount

mu    = mean(log(x))
sigma = sd(log(x))

# this is lognormal mean and median

mean_sev   = exp(mu + 0.5*sigma^2)
median_sev = exp(mu)

c(meanlog = mu, sdlog = sigma, mean_severity = mean_sev, median_severity = median_sev)

# notice the right skew of the data (mean to median)
```

## Visuals
```{r}

# A histogram of my log normal.
ggplot(data.frame(x), aes(x = log(x))) +
  geom_histogram(fill = "steelblue", color = "white") +
  geom_vline(xintercept = mu, color = "red", size=.8) + # Mean in log-space
  geom_vline(xintercept = median(log(x)), color = "green", size=.8) + # Mean in log-space
  labs(title = "Log-transformed Claim Severity",
       subtitle = "Red line represents the log-mean, Green represents the log median")


```


```{r}

ggplot(data.frame(x), aes(sample = log(x))) +
  stat_qq(color = "steelblue", alpha = 0.5) + 
  stat_qq_line(color = "red") +
  labs(title = "Normal Q-Q Plot (Log-transformed Claims)",
       x = "Theoretical Quantiles",
       y = "Sample Quantiles")
```

## Comapring Distribution to Gamma

```{r}


# Lognormal log-likelihood + AIC

loglik_logn = sum(dlnorm(x, meanlog = mu, sdlog = sigma, log = TRUE))
aic_logn = 2*2 - 2*loglik_logn

# Gamma MOM estimates + log-likelihood + AIC 
m = mean(x); v = var(x)
shape_hat = m^2 / v
rate_hat  = m / v

loglik_gam = sum(dgamma(x, shape = shape_hat, rate = rate_hat, log = TRUE))
aic_gam = 2*2 - 2*loglik_gam

data.frame(
  Model = c("Lognormal", "Gamma"),
  AIC   = c(aic_logn, aic_gam),
  row.names = NULL
)

# we will choose lognormal, better fit then gamma

```

## Tail Summaries

```{r}

q95 = qlnorm(0.95, meanlog = mu, sdlog = sigma)
q99 = qlnorm(0.99, meanlog = mu, sdlog = sigma)

pander(c(q95_logn = q95, q99_logn = q99))

pander(quantile(x, probs = c(0.90, 0.95, 0.99)))

pander(mean(x))

```

## Final report Sumamry
```{r}

# chat-gpt

delta_aic = aic_gam - aic_logn

cat(
  "- **Baseline frequency (lambda_hat):** ", round(lambda_hat, 4), " claims per exposure-year\n",
  "- **Overdispersion (Var/Mean):** ", round(dispersion_ratio, 3), " → Negative Binomial preferred\n",
  "- **Severity model:** Lognormal (meanlog = ", round(mu, 3), ", sdlog = ", round(sigma, 3), ")\n",
  "- **Mean severity (Lognormal):** $", format(round(mean_sev, 0), big.mark=","), "\n",
  "- **AIC comparison (Gamma - Lognormal):** ΔAIC = ", round(delta_aic, 1), " (positive favors Lognormal)\n",
  sep = ""
)

```









## Part 2

```{r}

# r = theta, p = tehta /(theta + mu)
fit_nb = glm.nb(claims_pol ~ 1 + offset(log(exposure_pol)), data = work_comp_pol)
summary(fit_nb)

# intercept
lambda_hat_nb = exp(coef(fit_nb)[1])  

# r 
theta_nb = fit_nb$theta

c(lambda_hat_nb = lambda_hat_nb, theta_nb = theta_nb)

```

## Time to use your severity and freq distributions to run some simulations!! We are using these simulations to see what the expected losses will be on average in the worst stress years as well as the worst 1% of years. (This is what Var and TVaR is)

```{r}

# total exposure
Exposure_period = sum(work_comp_pol$exposure_pol)

# r times the total exposure, this is akin finding the rate for the entire company, instead of one policy
mu_period = as.numeric(lambda_hat_nb) * Exposure_period
mu_period

# this is the rate for the entire company, multiplied by the mean claim cost
expected_aggregate = as.numeric(lambda_hat_nb) * Exposure_period * mean_sev
expected_aggregate

# for reproducing it
set.seed(42)
n_sims = 30000

# total losses
agg_losses = numeric(n_sims)

# run 30k times
for (i in seq_len(n_sims)) {
  # this creates a vector of frequencies (similating data) ie: 34 calims this year 47 nect
  N = rnbinom(1, size = theta_nb, mu = mu_period)
  if (N == 0) {
    agg_losses[i] = 0
  } else {
    # makes the severity, ie a cost for a claim
    agg_losses[i] = sum(rlnorm(N, meanlog = mu, sdlog = sigma))
  }
}

# average of all 30k years of claims
mean_L  = mean(agg_losses)
# value at risk (at what cost does the worst 99% start at)
VaR_99  = as.numeric(quantile(agg_losses, 0.99, names = FALSE))
# mean value at risk (average cost of the worst 1%)
TVaR_99 = mean(agg_losses[agg_losses >= VaR_99])

c(sim_mean = mean_L, theo_mean = expected_aggregate, ratio = mean_L/expected_aggregate,
  VaR_99 = VaR_99, TVaR_99 = TVaR_99)

```

## Fitting models/simulation for each solar system

```{r}
# Clean the solar_system names in both datasets
work_comp_freq$solar_system_clean = sub("_.*", "", work_comp_freq$solar_system)
work_comp_sev$solar_system_clean  = sub("_.*", "", work_comp_sev$solar_system)

# Verify the cleaning - should only see Epsilon, Zeta, Helionis Cluster, and maybe NA
unique(work_comp_freq$solar_system_clean)
```

```{r}

# Helper function to get the median for each column inputted, for a specific cleaned planet name cause they gave us crappy data there
get_planet_med = function(planet_name, col_name) {
  val = work_comp_freq %>% 
    filter(solar_system_clean == planet_name) %>% 
    pull(col_name)
  return(median(val, na.rm = TRUE))
}

get_perc = function(col_name, perc) {
  val = work_comp_freq %>% 
    pull(col_name)
  return(quantile(val, probs = perc, na.rm = TRUE))
}

profiles = rbind(
  Bayesia = data.frame(
    gravity_level           = get_perc("gravity_level", .85),
    psych_stress_index      = get_perc("psych_stress_index", .75),
    experience_yrs          = get_perc("experience_yrs", .6),
    safety_training_index   = get_perc("safety_training_index", .7),
    hours_per_week          = get_perc("hours_per_week", .5),
    protective_gear_quality = get_perc("protective_gear_quality", .9),
    exposure_pol            = 1
  ),
  Orion = data.frame(
    gravity_level           = get_perc("gravity_level", .45),
    psych_stress_index      = get_perc("psych_stress_index", .8),
    experience_yrs          = get_perc("experience_yrs", .3),
    safety_training_index   = get_perc("safety_training_index", .5),
    hours_per_week          = get_perc("hours_per_week", .65),
    protective_gear_quality = get_perc("protective_gear_quality", .6),
    exposure_pol            = 1
  ),
  Helionis_Cluster = data.frame(
    gravity_level           = get_planet_med("Helionis Cluster", "gravity_level"),
    psych_stress_index      = get_planet_med("Helionis Cluster", "psych_stress_index"),
    experience_yrs          = get_planet_med("Helionis Cluster", "experience_yrs"),
    safety_training_index   = get_planet_med("Helionis Cluster", "safety_training_index"),
    hours_per_week          = get_planet_med("Helionis Cluster", "hours_per_week"),
    protective_gear_quality = get_planet_med("Helionis Cluster", "protective_gear_quality"),
    exposure_pol            = 1
  )
)

# Baseline Profile (Overall Median), this will give us how to compare the overal middle to each levels planetary middle value
baseline = data.frame(
  gravity_level           = median(work_comp_freq$gravity_level, na.rm=TRUE),
  psych_stress_index      = median(work_comp_freq$psych_stress_index, na.rm=TRUE),
  experience_yrs          = median(work_comp_freq$experience_yrs, na.rm=TRUE),
  safety_training_index   = median(work_comp_freq$safety_training_index, na.rm=TRUE),
  hours_per_week          = median(work_comp_freq$hours_per_week, na.rm=TRUE),
  protective_gear_quality = median(work_comp_freq$protective_gear_quality, na.rm=TRUE),
  exposure_pol            = 1
)
```

```{r}
# Does the freq of claims change (noooo)
freq_base = as.numeric(predict(fit_nb, newdata = baseline, type = "response"))
freq_sys  = as.numeric(predict(fit_nb, newdata = profiles, type = "response"))
freq_mult = freq_sys / freq_base
names(freq_mult) = rownames(profiles)

# Severity Model - This is a log normal regression to predict log claim amounts on said factors
fit_sev = glm(log(claim_amount) ~ gravity_level + psych_stress_index + 
               experience_yrs + safety_training_index + hours_per_week + 
               protective_gear_quality + occupation, 
               data = work_comp_sev)
summary(fit_sev)

# Find common occupation
mode_occ = names(sort(table(work_comp_sev$occupation), decreasing = TRUE))[1]
baseline_sev = transform(baseline, occupation = mode_occ)
profiles_sev = cbind(profiles, occupation = mode_occ)

# Severity Multipliers
m_base = as.numeric(predict(fit_sev, newdata = baseline_sev))
m_sys  = as.numeric(predict(fit_sev, newdata = profiles_sev))
sev_mult = exp(m_sys - m_base)
names(sev_mult) = rownames(profiles)
```

```{r}
simulate_system = function(E_period, f_mult, s_mult, n_sims = 30000, seed = 42) {
  set.seed(seed)
  
  # Return NAs if multipliers are broken
  if(any(is.na(c(f_mult, s_mult)))) return(c(Expected_Loss = NA, VaR_99 = NA, TVaR_99 = NA))
  
  mu_period_adj = as.numeric(lambda_hat_nb) * f_mult * E_period
  
  agg = numeric(n_sims)
  for (i in seq_len(n_sims)) {
    N = rnbinom(1, size = theta_nb, mu = mu_period_adj)
    
    if (N == 0) {
      agg[i] = 0
    } else {
      # Use log-normal parameters from your previous step
      vals = rlnorm(N, meanlog = mu + log(s_mult), sdlog = sigma)
      # Capping the extreme values
      vals[is.infinite(vals)] = 1e8
      agg[i] = sum(vals)
    }
  }
  
  # Calculate value at ris kmetrics with na.rm to ensure stability
  VaR_99  = as.numeric(quantile(agg, 0.99, na.rm = TRUE, names = FALSE))
  TVaR_99 = if(is.na(VaR_99)) NA else mean(agg[agg >= VaR_99], na.rm = TRUE)
  
  c(Expected_Loss = mean(agg, na.rm = TRUE), VaR_99 = VaR_99, TVaR_99 = TVaR_99)
}

# Run simulation for all three planets solar systems or whateer they are
E_total = sum(work_comp_pol$exposure_pol)

out = rbind(
  Bayesia         = simulate_system(E_total, freq_mult["Bayesia"],          sev_mult["Bayesia"]),
  Orion             = simulate_system(E_total, freq_mult["Orion"],             sev_mult["Orion"]),
  Helionis_Cluster = simulate_system(E_total, freq_mult["Helionis_Cluster"], sev_mult["Helionis_Cluster"])
)

# This essentially displays the VaR and TVaR measures for each solar system in regards to my claim..
print(out)
```



## This is for each existing planetary system, a between groups analysis 

```{r}
# Building profiles based on the three given solar systems, we are doing the median to capture the middle of each level
profiles = rbind(
  Epsilon = data.frame(
    gravity_level           = get_planet_med("Epsilon", "gravity_level"),
    psych_stress_index      = get_planet_med("Epsilon", "psych_stress_index"),
    experience_yrs          = get_planet_med("Epsilon", "experience_yrs"),
    safety_training_index   = get_planet_med("Epsilon", "safety_training_index"),
    hours_per_week          = get_planet_med("Epsilon", "hours_per_week"),
    protective_gear_quality = get_planet_med("Epsilon", "protective_gear_quality"),
    exposure_pol            = 1
  ),
  Zeta = data.frame(
    gravity_level           = get_planet_med("Zeta", "gravity_level"),
    psych_stress_index      = get_planet_med("Zeta", "psych_stress_index"),
    experience_yrs          = get_planet_med("Zeta", "experience_yrs"),
    safety_training_index   = get_planet_med("Zeta", "safety_training_index"),
    hours_per_week          = get_planet_med("Zeta", "hours_per_week"),
    protective_gear_quality = get_planet_med("Zeta", "protective_gear_quality"),
    exposure_pol            = 1
  ),
  Helionis_Cluster = data.frame(
    gravity_level           = get_planet_med("Helionis Cluster", "gravity_level"),
    psych_stress_index      = get_planet_med("Helionis Cluster", "psych_stress_index"),
    experience_yrs          = get_planet_med("Helionis Cluster", "experience_yrs"),
    safety_training_index   = get_planet_med("Helionis Cluster", "safety_training_index"),
    hours_per_week          = get_planet_med("Helionis Cluster", "hours_per_week"),
    protective_gear_quality = get_planet_med("Helionis Cluster", "protective_gear_quality"),
    exposure_pol            = 1
  )
)

# Baseline Profile (Overall Median), this will give us how to compare the overal middle to each levels planetary middle value
baseline = data.frame(
  gravity_level           = median(work_comp_freq$gravity_level, na.rm=TRUE),
  psych_stress_index      = median(work_comp_freq$psych_stress_index, na.rm=TRUE),
  experience_yrs          = median(work_comp_freq$experience_yrs, na.rm=TRUE),
  safety_training_index   = median(work_comp_freq$safety_training_index, na.rm=TRUE),
  hours_per_week          = median(work_comp_freq$hours_per_week, na.rm=TRUE),
  protective_gear_quality = median(work_comp_freq$protective_gear_quality, na.rm=TRUE),
  exposure_pol            = 1
)
```

```{r}
# Does the frequency of claims change (no, we want it to be = for analysis )
freq_base = as.numeric(predict(fit_nb, newdata = baseline, type = "response"))
freq_sys  = as.numeric(predict(fit_nb, newdata = profiles, type = "response"))
freq_mult = freq_sys / freq_base
names(freq_mult) = rownames(profiles)

# Severity Model - This is a log normal regression to predict log claim amounts on said factors
fit_sev = glm(log(claim_amount) ~ gravity_level + psych_stress_index + 
               experience_yrs + safety_training_index + hours_per_week + 
               protective_gear_quality + occupation, 
               data = work_comp_sev)
summary(fit_sev)

# Find common occupation
mode_occ = names(sort(table(work_comp_sev$occupation), decreasing = TRUE))[1]
baseline_sev = transform(baseline, occupation = mode_occ)
profiles_sev = cbind(profiles, occupation = mode_occ)

# Severity Multipliers
m_base = as.numeric(predict(fit_sev, newdata = baseline_sev))
m_sys  = as.numeric(predict(fit_sev, newdata = profiles_sev))
sev_mult = exp(m_sys - m_base)
names(sev_mult) = rownames(profiles)
```

```{r}
simulate_system = function(E_period, f_mult, s_mult, n_sims = 30000, seed = 42) {
  set.seed(seed)
  
  # Return NAs if multipliers are broken
  if(any(is.na(c(f_mult, s_mult)))) return(c(Expected_Loss = NA, VaR_99 = NA, TVaR_99 = NA))
  
  mu_period_adj = as.numeric(lambda_hat_nb) * f_mult * E_period
  
  agg = numeric(n_sims)
  for (i in seq_len(n_sims)) {
    N = rnbinom(1, size = theta_nb, mu = mu_period_adj)
    
    if (N == 0) {
      agg[i] = 0
    } else {
      # Use log-normal parameters from your previous step
      vals = rlnorm(N, meanlog = mu + log(s_mult), sdlog = sigma)
      # Capping the extreme values
      vals[is.infinite(vals)] = 1e8
      agg[i] = sum(vals)
    }
  }
  
  # Calculate value at ris kmetrics with na.rm to ensure stability
  VaR_99  = as.numeric(quantile(agg, 0.99, na.rm = TRUE, names = FALSE))
  TVaR_99 = if(is.na(VaR_99)) NA else mean(agg[agg >= VaR_99], na.rm = TRUE)
  
  c(Expected_Loss = mean(agg, na.rm = TRUE), VaR_99 = VaR_99, TVaR_99 = TVaR_99)
}

# Run simulation for all three planets solar systems or whateer they are
E_total = sum(work_comp_pol$exposure_pol)

out = rbind(
  Epsilon          = simulate_system(E_total, freq_mult["Epsilon"],          sev_mult["Epsilon"]),
  Zeta             = simulate_system(E_total, freq_mult["Zeta"],             sev_mult["Zeta"]),
  Helionis_Cluster = simulate_system(E_total, freq_mult["Helionis_Cluster"], sev_mult["Helionis_Cluster"])
)

# This essentially displays the VaR and TVaR measures for each solar system in regards to my claim..
print(out)
```