---
title: "Workers Compensation"
format: html
editor_options: 
  chunk_output_type: console
---

# Load Data & Libraries

```{r}
pacman::p_load(tidyverse, readxl, ggrepel, directlabels, readr, rio, knitr, 
               pander, downloader, dplyr, plotly, ggplot2, ggthemes, haven, 
               ggdist, janitor, MASS, here)

path = "srcsc-2026-claims-workers-comp.xlsx"

work_comp_freq = read_excel(path, sheet = "freq")
work_comp_sev  = read_excel(path, sheet = "sev")
```

---

# Part 1: Finding the Right Distributions

### Combining Rows and Filtering

```{r}
# Aggregate to policy level - one row per policy
work_comp_pol = work_comp_freq %>%
  filter(!is.na(exposure), !is.na(claim_count), exposure > 0) %>%
  group_by(policy_id) %>%
  summarise(
    exposure_pol = sum(exposure, na.rm = TRUE),
    claims_pol   = sum(claim_count, na.rm = TRUE),
    .groups = "drop"
  )
```

### Estimating Claim Frequency Distribution

```{r}
# Exposure-weighted claim rate
lambda_hat = sum(work_comp_pol$claims_pol) / sum(work_comp_pol$exposure_pol)
lambda_hat

# Check for overdispersion - if variance >> mean, Poisson is not appropriate
mean_claims      = mean(work_comp_pol$claims_pol)
var_claims       = var(work_comp_pol$claims_pol)
dispersion_ratio = var_claims / mean_claims

c(mean_claims = mean_claims, var_claims = var_claims, dispersion_ratio = dispersion_ratio)

# Dispersion > 1 means variance exceeds mean then Negative Binomial preferred over Poisson
```

### Estimating Claim Severity Distribution

```{r}
# Filter claim amounts to valid positives
x = work_comp_sev$claim_amount
x = as.numeric(x)
x = x[is.finite(x) & !is.na(x) & x > 0]

# Raw lognormal parameters (used only for initial exploration and AIC comparison)
mu_raw    = mean(log(x))
sigma_raw = sd(log(x))

mean_sev_raw   = exp(mu_raw + 0.5 * sigma_raw^2)
median_sev_raw = exp(mu_raw)

c(meanlog = mu_raw, sdlog = sigma_raw, 
  mean_severity = mean_sev_raw, median_severity = median_sev_raw)
# Note the right skew: mean >> median, typical of insurance claim data
```

### Severity Visuals

```{r}
# Log-transformed histogram - should look approximately normal if lognormal fits
ggplot(data.frame(x), aes(x = log(x))) +
  geom_histogram(fill = "steelblue", color = "white", bins = 40) +
  geom_vline(xintercept = mu_raw, color = "red", linewidth = 0.8) +
  geom_vline(xintercept = median(log(x)), color = "green", linewidth = 0.8) +
  labs(title = "Log-transformed Claim Severity",
       subtitle = "Red = log-mean, Green = log-median",
       x = "log(Claim Amount)", y = "Count")
```

```{r}
# Q-Q plot - points should follow the line if log-normal is a good fit
ggplot(data.frame(x), aes(sample = log(x))) +
  stat_qq(color = "steelblue", alpha = 0.5) +
  stat_qq_line(color = "red") +
  labs(title = "Normal Q-Q Plot (Log-transformed Claims)",
       x = "Theoretical Quantiles", y = "Sample Quantiles")
```

### Comparing Severity Distributions: Lognormal vs Gamma

```{r}
# Lognormal AIC
loglik_logn = sum(dlnorm(x, meanlog = mu_raw, sdlog = sigma_raw, log = TRUE))
aic_logn    = 2*2 - 2*loglik_logn

# Gamma MOM estimates + AIC
m         = mean(x); v = var(x)
shape_hat = m^2 / v
rate_hat  = m / v

loglik_gam = sum(dgamma(x, shape = shape_hat, rate = rate_hat, log = TRUE))
aic_gam    = 2*2 - 2*loglik_gam

delta_aic = aic_gam - aic_logn

data.frame(
  Model = c("Lognormal", "Gamma"),
  AIC   = c(aic_logn, aic_gam),
  row.names = NULL
)
# Positive delta AIC = Lognormal fits better → proceed with Lognormal for severity
```

### Tail Summaries

```{r}
q95 = qlnorm(0.95, meanlog = mu_raw, sdlog = sigma_raw)
q99 = qlnorm(0.99, meanlog = mu_raw, sdlog = sigma_raw)

pander(c(q95_logn = q95, q99_logn = q99))
pander(quantile(x, probs = c(0.90, 0.95, 0.99)))
pander(mean(x))
```

### Part 1 Summary

```{r}
cat(
  "- **Baseline frequency (lambda_hat):** ", round(lambda_hat, 4), " claims per exposure-year\n",
  "- **Overdispersion (Var/Mean):** ", round(dispersion_ratio, 3), " → Negative Binomial preferred\n",
  "- **Severity model:** Lognormal (raw: meanlog = ", round(mu_raw, 3), ", sdlog = ", round(sigma_raw, 3), ")\n",
  "- **Mean severity (raw Lognormal):** $", format(round(mean_sev_raw, 0), big.mark=","), "\n",
  "- **AIC comparison (Gamma - Lognormal):** ΔAIC = ", round(delta_aic, 1), " → Lognormal preferred\n",
  sep = ""
)
```

---

# Part 2: Fitting the Models

### Fitting Frequency Model (Negative Binomial GLM)

```{r}
# NegBin GLM with log-exposure offset
# theta = dispersion (size), lambda = rate per unit exposure
fit_nb = glm.nb(claims_pol ~ 1 + offset(log(exposure_pol)), data = work_comp_pol)
summary(fit_nb)

lambda_hat_nb = exp(coef(fit_nb)[1])   # claim rate per exposure unit
theta_nb      = fit_nb$theta           # NegBin dispersion parameter

c(lambda_hat_nb = lambda_hat_nb, theta_nb = theta_nb)
```

```{r}
# Scale up to entire portfolio
Exposure_period  = sum(work_comp_pol$exposure_pol)
mu_period        = as.numeric(lambda_hat_nb) * Exposure_period  # expected claims, whole book

c(Exposure_period = Exposure_period, mu_period = mu_period)
```

### Cleaning Occupation Categories

```{r}
# Collapse garbled occupation levels into "Other" to avoid unstable coefficients
work_comp_sev$occupation_clean  = ifelse(grepl("\\?\\?\\?", work_comp_sev$occupation),
                                          "Other", as.character(work_comp_sev$occupation))

work_comp_freq$occupation_clean = ifelse(grepl("\\?\\?\\?", work_comp_freq$occupation),
                                          "Other", as.character(work_comp_freq$occupation))
```

### Fitting Severity Model (Lognormal Regression)

```{r}
# Log-normal regression: only include significant predictors for pricing stability
# Removed: gravity_level, experience_yrs, safety_training_index, hours_per_week (all p > 0.3)
fit_sev = glm(log(claim_amount) ~ psych_stress_index + protective_gear_quality + occupation_clean,
              data = work_comp_sev)

summary(fit_sev)

# mu and sigma now come entirely from fit_sev - fully consistent for pricing
# mu    = predicted log-mean at baseline (derived below per profile)
# sigma = residual standard deviation after covariates are accounted for
sigma = sqrt(summary(fit_sev)$dispersion)
cat("sigma (from regression):", sigma, "\n")
```

### Solar System Name Cleaning

```{r}
work_comp_freq$solar_system_clean = sub("_.*", "", work_comp_freq$solar_system)
work_comp_sev$solar_system_clean  = sub("_.*", "", work_comp_sev$solar_system)

unique(work_comp_freq$solar_system_clean)
```

---

# Part 3: Building Territory Profiles & Multipliers

### Profile Construction

```{r}
# Helper: median covariate for a specific solar system (used for Helionis which exists in data)
get_planet_med = function(planet_name, col_name) {
  val = work_comp_freq %>%
    filter(solar_system_clean == planet_name) %>%
    pull(col_name)
  return(median(val, na.rm = TRUE))
}

# Helper: global percentile (used for Bayesia and Orion which do NOT exist in data)
get_perc = function(col_name, perc) {
  val = work_comp_freq %>% pull(col_name)
  return(quantile(val, probs = perc, na.rm = TRUE))
}

# Profiles - DO NOT CHANGE these percentile assignments
profiles = rbind(
  Bayesia = data.frame(
    psych_stress_index      = get_perc("psych_stress_index", .75),
    experience_yrs          = get_perc("experience_yrs", .6),
    safety_training_index   = get_perc("safety_training_index", .7),
    hours_per_week          = get_perc("hours_per_week", .5),
    protective_gear_quality = get_perc("protective_gear_quality", .9),
    gravity_level           = get_perc("gravity_level", .85),
    exposure_pol            = 1
  ),
  Orion = data.frame(
    psych_stress_index      = get_perc("psych_stress_index", .8),
    experience_yrs          = get_perc("experience_yrs", .3),
    safety_training_index   = get_perc("safety_training_index", .5),
    hours_per_week          = get_perc("hours_per_week", .65),
    protective_gear_quality = get_perc("protective_gear_quality", .6),
    gravity_level           = get_perc("gravity_level", .45),
    exposure_pol            = 1
  ),
  Helionis_Cluster = data.frame(
    psych_stress_index      = get_planet_med("Helionis Cluster", "psych_stress_index"),
    experience_yrs          = get_planet_med("Helionis Cluster", "experience_yrs"),
    safety_training_index   = get_planet_med("Helionis Cluster", "safety_training_index"),
    hours_per_week          = get_planet_med("Helionis Cluster", "hours_per_week"),
    protective_gear_quality = get_planet_med("Helionis Cluster", "protective_gear_quality"),
    gravity_level           = get_planet_med("Helionis Cluster", "gravity_level"),
    exposure_pol            = 1
  )
)

# Overall median baseline - the reference point for all multipliers
baseline = data.frame(
  psych_stress_index      = median(work_comp_freq$psych_stress_index, na.rm = TRUE),
  experience_yrs          = median(work_comp_freq$experience_yrs, na.rm = TRUE),
  safety_training_index   = median(work_comp_freq$safety_training_index, na.rm = TRUE),
  hours_per_week          = median(work_comp_freq$hours_per_week, na.rm = TRUE),
  protective_gear_quality = median(work_comp_freq$protective_gear_quality, na.rm = TRUE),
  gravity_level           = median(work_comp_freq$gravity_level, na.rm = TRUE),
  exposure_pol            = 1
)
```

### Frequency Multipliers

```{r}
# NegBin GLM frequency prediction - all profiles return same rate since no covariates in fit_nb
# (frequency model only uses intercept + exposure offset)
freq_base = as.numeric(predict(fit_nb, newdata = baseline, type = "response"))
freq_sys  = as.numeric(predict(fit_nb, newdata = profiles, type = "response"))
freq_mult = freq_sys / freq_base
names(freq_mult) = rownames(profiles)

cat("Frequency multipliers (all ~1.0 - expected since fit_nb has no covariates):\n")
print(freq_mult)
```

### Severity Multipliers

```{r}
# Most common (clean) occupation used as the reference level
mode_occ     = names(sort(table(work_comp_sev$occupation_clean), decreasing = TRUE))[1]
baseline_sev = transform(baseline, occupation_clean = mode_occ)
profiles_sev = cbind(profiles, occupation_clean = mode_occ)

# mu (log-mean) at baseline - this is now consistent with fit_sev
mu = as.numeric(predict(fit_sev, newdata = baseline_sev))

# Predicted log-mean at each profile
m_sys    = as.numeric(predict(fit_sev, newdata = profiles_sev))

# Multiplicative severity relativities vs baseline
sev_mult = exp(m_sys - mu)
names(sev_mult) = rownames(profiles)

cat("Baseline mu (log-mean from regression):", round(mu, 4), "\n")
cat("Sigma (residual sd from regression):   ", round(sigma, 4), "\n\n")
cat("Severity multipliers:\n")
print(sev_mult)

# Implied mean severity per territory
mean_sev_baseline = exp(mu + 0.5 * sigma^2)
cat("\nBaseline mean severity: $", format(round(mean_sev_baseline, 0), big.mark=","), "\n")
cat("Expected aggregate (baseline): $", 
    format(round(lambda_hat_nb * Exposure_period * mean_sev_baseline, 0), big.mark=","), "\n")
```

---

# Part 4: Monte Carlo Simulation by Territory

### Simulation Function

```{r}
simulate_system = function(E_period, f_mult, s_mult, mu_base, sigma, 
                            n_sims = 30000, seed = 42) {
  set.seed(seed)
  
  if (any(is.na(c(f_mult, s_mult)))) {
    return(c(Expected_Loss = NA, Std_Dev = NA, Variance = NA, VaR_99 = NA, TVaR_99 = NA))
  }
  
  # Adjusted expected claim count for this territory
  mu_period_adj = as.numeric(lambda_hat_nb) * f_mult * E_period
  
  agg = numeric(n_sims)
  
  for (i in seq_len(n_sims)) {
    # Frequency: NegBin with fitted theta and territory-adjusted mu
    N = rnbinom(1, size = theta_nb, mu = mu_period_adj)
    
    if (N == 0) {
      agg[i] = 0
    } else {
      # Severity: LogNormal with regression-derived mu and sigma
      # s_mult shifts the log-mean multiplicatively (log scale addition)
      vals = rlnorm(N, meanlog = mu_base + log(s_mult), sdlog = sigma)
      vals[is.infinite(vals)] = 1e8   # cap extreme outliers for stability
      agg[i] = sum(vals)
    }
  }
  
  # Risk metrics
  sim_mean = mean(agg, na.rm = TRUE)
  sim_var  = var(agg, na.rm = TRUE)
  sim_sd   = sd(agg, na.rm = TRUE)
  VaR_99   = as.numeric(quantile(agg, 0.99, na.rm = TRUE, names = FALSE))
  TVaR_99  = if (is.na(VaR_99)) NA else mean(agg[agg >= VaR_99], na.rm = TRUE)
  
  c(Expected_Loss = sim_mean, 
    Std_Dev       = sim_sd,
    Variance      = sim_var,
    VaR_99        = VaR_99, 
    TVaR_99       = TVaR_99)
}
```

### Run Simulations

```{r}
E_total = sum(work_comp_pol$exposure_pol)

# Each territory gets a unique seed so draws are independent
out = rbind(
  Bayesia          = simulate_system(E_total, freq_mult["Bayesia"],
                                     sev_mult["Bayesia"],
                                     mu_base = mu, sigma = sigma, seed = 42),
  Orion            = simulate_system(E_total, freq_mult["Orion"],
                                     sev_mult["Orion"],
                                     mu_base = mu, sigma = sigma, seed = 43),
  Helionis_Cluster = simulate_system(E_total, freq_mult["Helionis_Cluster"],
                                     sev_mult["Helionis_Cluster"],
                                     mu_base = mu, sigma = sigma, seed = 44)
)

print(round(out, 0))
```

### Territory Comparison Table

```{r}
# Format for readability
out_fmt = as.data.frame(out)
out_fmt[] = lapply(out_fmt, function(col) paste0("$", format(round(col, 0), big.mark = ",")))

knitr::kable(out_fmt, 
             caption = "Aggregate Loss Projections by Territory (30,000 Simulations)",
             col.names = c("Expected Loss", "Std Dev", "Variance", "VaR (99%)", "TVaR (99%)"))
```

### Severity & Frequency Multiplier Summary

```{r}
mult_summary = data.frame(
  Territory        = names(freq_mult),
  Freq_Multiplier  = round(as.numeric(freq_mult), 4),
  Sev_Multiplier   = round(as.numeric(sev_mult), 4),
  Combined_Mult    = round(as.numeric(freq_mult) * as.numeric(sev_mult), 4)
)

knitr::kable(mult_summary, caption = "Frequency, Severity, and Combined Multipliers vs Baseline")
```

### Visualization: Expected Loss Distribution by Territory

```{r}
# Re-run sims and store all draws for plotting
set.seed(42)
sim_draws = lapply(1:3, function(i) {
  sys    = rownames(profiles)[i]
  f_mult = freq_mult[sys]
  s_mult = sev_mult[sys]
  seed_i = 41 + i
  set.seed(seed_i)
  
  mu_period_adj = as.numeric(lambda_hat_nb) * f_mult * E_total
  agg = numeric(10000)
  for (j in seq_len(10000)) {
    N = rnbinom(1, size = theta_nb, mu = mu_period_adj)
    if (N == 0) { agg[j] = 0 } 
    else {
      vals = rlnorm(N, meanlog = mu + log(s_mult), sdlog = sigma)
      vals[is.infinite(vals)] = 1e8
      agg[j] = sum(vals)
    }
  }
  data.frame(Territory = sys, Loss = agg)
})

sim_df = bind_rows(sim_draws)

ggplot(sim_df, aes(x = Loss / 1e6, fill = Territory)) +
  geom_density(alpha = 0.5) +
  labs(title = "Simulated Aggregate Loss Distribution by Territory",
       x = "Aggregate Loss ($ Millions)", y = "Density") +
  theme_minimal()
```



# EDA for pricing!!!

```{r}
ggplot(work_comp_sev, aes(x = protective_gear_quality, y = claim_amount)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm") +
  scale_y_log10() +
  labs(title = "Claim Severity vs Protective Gear Quality",
       subtitle = "Negative relationship justifies compliance credit")

```


```{r}
ggplot(work_comp_sev, aes(x = psych_stress_index, y = claim_amount)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm") +
  scale_y_log10() +
  labs(title = "Claim Severity vs Psych Stress Index",
       subtitle = "Positive relationship justifies waiting period")

```


```{r}
ggplot(work_comp_sev, aes(x = reorder(occupation_clean, claim_amount, median), 
                           y = claim_amount)) +
  geom_boxplot(fill = "steelblue", alpha = 0.6) +
  scale_y_log10() +
  coord_flip() +
  labs(title = "Claim Severity by Occupation",
       subtitle = "Justifies high-severity occupation sublimit")

```


```{r}
out_df = as.data.frame(out) %>%
  rownames_to_column("System")

ggplot(out_df, aes(x = System, y = TVaR_99 / 1e6, fill = System)) +
  geom_col() +
  labs(title = "TVaR (99%) by Solar System",
       y = "TVaR in $ Millions",
       subtitle = "Drives system-specific endorsement design")
```


```{r}
# Shows exactly how many claims would be capped by a sublimit
sublimit = 2 * median(work_comp_sev$claim_amount)

ggplot(work_comp_sev, aes(x = claim_amount)) +
  geom_histogram(fill = "steelblue", color = "white", bins = 50) +
  geom_vline(xintercept = sublimit, color = "red", linewidth = 1, linetype = "dashed") +
  scale_x_log10(labels = scales::comma) +
  annotate("text", x = sublimit * 1.3, y = 100, 
           label = paste0("Sublimit: $", format(round(sublimit,0), big.mark=",")), 
           color = "red", hjust = 0) +
  labs(title = "Claim Severity Distribution with Proposed Sublimit",
       subtitle = "Claims to the right of the line would be capped",
       x = "Claim Amount (log scale)", y = "Count")

```


```{r}
# Shows which occupations are driving the excess claims
work_comp_sev %>%
  mutate(excess = pmax(claim_amount - sublimit, 0)) %>%
  group_by(occupation_clean) %>%
  summarise(total_excess = sum(excess), avg_excess = mean(excess), .groups = "drop") %>%
  filter(total_excess > 0) %>%
  ggplot(aes(x = reorder(occupation_clean, total_excess), y = total_excess / 1e6)) +
  geom_col(fill = "tomato", alpha = 0.8) +
  coord_flip() +
  labs(title = "Total Excess Loss Above Sublimit by Occupation",
       subtitle = "Justifies occupation-specific sublimit endorsement",
       x = NULL, y = "Excess Loss ($ Millions)")

```



```{r}
# Shows the threshold where a compliance credit  would kick in
gear_threshold = quantile(work_comp_sev$protective_gear_quality, 0.80)

ggplot(work_comp_sev, aes(x = protective_gear_quality, y = claim_amount, 
                           color = protective_gear_quality >= gear_threshold)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  geom_vline(xintercept = gear_threshold, linetype = "dashed", color = "darkgreen", linewidth = 1) +
  scale_y_log10(labels = scales::comma) +
  scale_color_manual(values = c("FALSE" = "tomato", "TRUE" = "steelblue"),
                     labels = c("Below Threshold", "Qualifies for Credit")) +
  annotate("text", x = gear_threshold + 0.02, y = max(work_comp_sev$claim_amount) * 0.8,
           label = "Credit Threshold (80th pct)", color = "darkgreen", hjust = 0) +
  labs(title = "Claim Severity vs Protective Gear Quality",
       subtitle = "Blue points qualify for compliance credit — clear severity reduction",
       x = "Protective Gear Quality Index", y = "Claim Amount (log scale)",
       color = NULL)

```


```{r}
# Merges freq and sev data to show where high freq AND high sev overlap
freq_summary = work_comp_freq %>%
  group_by(solar_system_clean) %>%
  summarise(avg_claims = mean(claim_count, na.rm = TRUE), .groups = "drop")

sev_summary = work_comp_sev %>%
  group_by(solar_system_clean) %>%
  summarise(avg_severity = mean(claim_amount, na.rm = TRUE), .groups = "drop")

left_join(freq_summary, sev_summary, by = "solar_system_clean") %>%
  filter(!is.na(solar_system_clean)) %>%
  ggplot(aes(x = avg_claims, y = avg_severity / 1e3, label = solar_system_clean)) +
  geom_point(size = 5, color = "steelblue") +
  ggrepel::geom_label_repel(size = 4) +
  labs(title = "Frequency vs Severity by Solar System",
       subtitle = "Upper-right quadrant = highest risk systems needing endorsements",
       x = "Average Claims per Policy", y = "Average Severity ($000s)")

```




# Pricing !!!!!

```{r}
# searching for a threshol for protective gear to create price
gear_threshold <- quantile(work_comp_sev$protective_gear_quality, 0.80)

ggplot(work_comp_sev, aes(x = protective_gear_quality, y = claim_amount)) +
  geom_point(aes(color = protective_gear_quality >= gear_threshold), alpha = 0.4) +
  geom_smooth(method = "lm", color = "black") +
  scale_y_log10(labels = scales::dollar) +
  scale_color_manual(values = c("darkred", "darkgreen"), 
                     labels = c("Standard Gear", "High-Quality Gear")) +
  labs(title = "Relationship: Gear Quality vs. Claim Severity",
       subtitle = "Proving the basis for a 'Safety Compliance Credit'",
       x = "Protective Gear Quality Index",
       y = "Claim Amount (Log Scale)",
       color = "Group") +
  theme_minimal()

# Correlation between Gear Quality and the Log of Claim Amount
cor_test <- cor.test(work_comp_sev$protective_gear_quality, log(work_comp_sev$claim_amount))

cat("Correlation Coefficient:", round(cor_test$estimate, 4), "\n")
cat("P-Value:", format.pval(cor_test$p.value), "\n")

```







```{r}
work_comp_sev_clean <- work_comp_sev %>%
  filter(!is.na(psych_stress_index), !is.na(claim_amount), claim_amount > 0)

stress_threshold <- quantile(work_comp_sev_clean$psych_stress_index, 0.75)

# Now run the plot and the correlation test
ggplot(work_comp_sev_clean, aes(x = psych_stress_index, y = claim_amount)) +
  geom_point(aes(color = psych_stress_index >= stress_threshold), alpha = 0.3) +
  geom_smooth(method = "lm", color = "darkred") +
  scale_y_log10(labels = scales::dollar) +
  scale_color_manual(values = c("gray50", "orange"), 
                     labels = c("Normal Stress", "High Stress Zone")) +
  labs(title = "Relationship: Psych Stress vs. Claim Severity",
       subtitle = "Justifying a 'Mental Health Risk' Adjustment",
       x = "Psych Stress Index",
       y = "Claim Amount (Log Scale)",
       color = "Category") +
  theme_minimal()

# Also add na.use to the correlation test to handle those NAs
cor_stress <- cor.test(work_comp_sev$psych_stress_index, log(work_comp_sev$claim_amount), 
                       use = "complete.obs")

cat("Stress Correlation Coefficient:", round(cor_stress$estimate, 4), "\n")
cat("P-Value:", format.pval(cor_stress$p.value), "\n")

```



```{r}
# Plotting occupation vs claim amount to find the 'expensive' outliers
ggplot(work_comp_sev_clean, aes(x = reorder(occupation_clean, claim_amount, median), 
                               y = claim_amount, fill = occupation_clean)) +
  geom_boxplot(alpha = 0.7, outlier.color = "red") +
  scale_y_log10(labels = scales::dollar) +
  coord_flip() +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(title = "Claim Severity by Occupation",
       subtitle = "Identifying High-Risk Groups for Sublimits",
       x = "Occupation",
       y = "Claim Amount (Log Scale)")


# Calculate the 95th percentile (Value at Risk) per occupation
occ_risk_summary <- work_comp_sev_clean %>%
  group_by(occupation_clean) %>%
  summarise(
    Median_Claim = median(claim_amount),
    nine5_Percentile = quantile(claim_amount, 0.95),
    Max_Claim = max(claim_amount),
    Sample_Size = n()
  ) %>%
  arrange(desc(nine5_Percentile))

print(occ_risk_summary)

```





```{r}
# Merge freq and sev data to see where high freq AND high sev overlap
freq_summary <- work_comp_freq %>%
  group_by(solar_system_clean) %>%
  summarise(avg_claims = mean(claim_count, na.rm = TRUE), .groups = "drop")

sev_summary <- work_comp_sev_clean %>%
  group_by(solar_system_clean) %>%
  summarise(avg_severity = mean(claim_amount, na.rm = TRUE), .groups = "drop")

territory_data <- left_join(freq_summary, sev_summary, by = "solar_system_clean") %>%
  filter(!is.na(solar_system_clean))

ggplot(territory_data, aes(x = avg_claims, y = avg_severity, label = solar_system_clean)) +
  geom_point(size = 5, color = "darkblue") +
  ggrepel::geom_label_repel() +
  scale_y_continuous(labels = scales::dollar) +
  labs(title = "Territory Risk Mapping: Frequency vs. Severity",
       subtitle = "Identifying Systems that require mandatory higher deductibles",
       x = "Average Claims per Policy", 
       y = "Average Claim Severity") +
  theme_minimal()



# Check coefficient of variation (Standard Deviation / Mean) by Territory
territory_risk <- work_comp_sev_clean %>%
  group_by(solar_system_clean) %>%
  summarise(
    Mean_Sev = mean(claim_amount),
    SD_Sev = sd(claim_amount),
    CV = SD_Sev / Mean_Sev,
    Total_Claims = n()
  ) %>%
  arrange(desc(CV))

print(territory_risk)
```



```{r}
# Experience vs Claim Severity
ggplot(work_comp_sev_clean, aes(x = experience_yrs, y = claim_amount)) +
  geom_point(alpha = 0.2, color = "darkgreen") +
  geom_smooth(method = "gam", color = "black") + # GAM shows non-linear trends better
  scale_y_log10(labels = scales::dollar) +
  labs(title = "Experience vs. Claim Severity",
       subtitle = "Proving the 'Learning Curve' for Experience Credits",
       x = "Years of Experience",
       y = "Claim Amount (Log Scale)") +
  theme_minimal()


# Categorize experience and check means
exp_groups <- work_comp_sev_clean %>%
  mutate(exp_tier = case_when(
    experience_yrs <= 2 ~ "Rookie (0-2)",
    experience_yrs > 2 & experience_yrs <= 10 ~ "Mid-Level (3-10)",
    experience_yrs > 10 ~ "Veteran (10+)"
  )) %>%
  group_by(exp_tier) %>%
  summarise(
    Mean_Severity = mean(claim_amount),
    Median_Severity = median(claim_amount),
    Max_Claim = max(claim_amount),
    Total_Claims = n()
  )

print(exp_groups)

```




```{r}
# agregate Severity Data to Policy Level first (Avoids many-to-many join)
# We take the mean of indices for policies with multiple claims
pol_sev_features <- work_comp_sev_clean %>%
  group_by(policy_id) %>%
  summarise(
    avg_gear = mean(protective_gear_quality, na.rm = TRUE),
    avg_stress = mean(psych_stress_index, na.rm = TRUE),
    # Use the first occupation listed for the policy
    occ_clean = first(occupation_clean),
    .groups = "drop"
  )

# Join and Price
work_comp_pricing_final <- work_comp_freq %>%
  left_join(pol_sev_features, by = "policy_id") %>%
  mutate(
    # Fill NAs using the global medians we calculated in Part 1/2
    avg_gear = replace_na(avg_gear, median(work_comp_sev_clean$protective_gear_quality, na.rm = TRUE)),
    avg_stress = replace_na(avg_stress, median(work_comp_sev_clean$psych_stress_index, na.rm = TRUE)),
    occ_clean = replace_na(occ_clean, "Other"),
    solar_system_clean = replace_na(solar_system_clean, "Unknown")
  ) %>%
  mutate(
    # ---  GEAR CREDIT (Proven by -0.1488 Cor) ---
    m1 = ifelse(avg_gear >= 4.0, 0.88, 1.00),
    
    # ---  STRESS SURCHARGE (Proven by 0.2441 Cor) ---
    m2 = ifelse(avg_stress >= 7.5, 1.15, 1.00),
    
    # ---  OCC SUBLIMIT (Proven by Executive/Manager Tail Risk) ---
    m3 = ifelse(occ_clean %in% c("Executive", "Manager"), 0.95, 1.00),
    
    # ---  TERRITORY (Proven by Helionis Cluster TVaR) ---
    m4 = ifelse(solar_system_clean == "Helionis", 1.23, 1.00),
    
    # --- EXPERIENCE (Proven by Mid-Level U-Shape) ---
    m5 = case_when(
      experience_yrs > 2 & experience_yrs <= 10 ~ 0.90,  # Mid-level discount
      experience_yrs <= 2 ~ 1.10,                        # Rookie surcharge
      TRUE ~ 1.05                                        # Veteran surcharge
    ),
    
    # Base Rate: $150 (Derived fromlambda_hat and expected loss)
    Base_Rate = 150.00,
    
    # FINAL PREMIUM CALCULATION
    Final_Premium = round(exposure * Base_Rate * m1 * m2 * m3 * m4 * m5, 2)
  )

# 3. Summary Table to verify it matches your Monte Carlo "out" values
pricing_summary <- work_comp_pricing_final %>%
  group_by(solar_system_clean) %>%
  summarise(
    Policies = n(),
    Total_Exposure = sum(exposure, na.rm = TRUE),
    Total_Premium = sum(Final_Premium, na.rm = TRUE),
    Avg_Premium = mean(Final_Premium / exposure, na.rm = TRUE)
  )

print(pricing_summary)

```


